Escape Analysis and Specialization in a Tracing JIT

### 2010-09-26 talk/pepm2011/paper.tex
more in the brackground section about tracing JITs

\authorinfo{Carl Friedrich Bolz \and Armin Rigo \and Antion Cuni \and Maciek Fijałkowski}
           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany}
           {cfbolz@gmx.de}

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. The PyPy
project is an environment for implementing dynamic programming languages. It's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient VM that also contains a just-in-time compiler. This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}

@@

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an interpreter for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way, that type inference becomes
possible. The language interpreter can thus be translated with the help of
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C.

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is not fully
automatic, but needs to be guided by the language implementer by some
source-code hints.

@@

traces and bridges

>>>>

\authorinfo{Carl Friedrich Bolz \and Antion Cuni \and Maciek Fijałkowski \and Samuele Pedroni \and Armin Rigo}
           {Heinrich-Heine-Universität Düsseldorf, STUPS Group, Germany XXX}
           {cfbolz@gmx.de XXX}

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT often takes the form of an
extension to an existing interpreter, which can be sped up that way. The PyPy
project is an environment for implementing dynamic programming languages. It's
approach to doing so is to straightforwardly implement an interpreter for the
to-be-implemented language, and then use powerful tools to turn the interpreter
into an efficient VM that also contains a just-in-time compiler. This compiler
is automatically generated from the interpreter using partial-evaluation-like
techniques \cite{bolz_tracing_2009}. The PyPy project and its approach to
tracing JIT compilers is described in Section~\ref{sec:Background}.

@@

The work described in this paper was done in the context of the PyPy project
\cite{armin_rigo_pypys_2006}. PyPy is an environment where dynamic languages can
be implemented in a simple yet efficient way. The approach taken when
implementing a language with PyPy is to write an interpreter for the language in
\emph{RPython} \cite{davide_ancona_rpython:_2007}. RPython ("restricted Python")
is a subset of Python chosen in such a way, that type inference becomes
possible. The language interpreter can thus be translated with the help of
PyPy's tools into a VM on the C level. Because the interpreter is written at a
relatively high level, the language implementation is kept free of low-level
details, such as object layout, garbage collection or memory model. Those
aspects of the final VM are woven into the generated code during the translation
to C. XXX languages that are done using PyPy

The feature that makes PyPy more than a compiler with a runtime system is it's
support for automated JIT compiler generation \cite{bolz_tracing_2009}. During
the translation to C, PyPy's tools can generate a just-in-time compiler for the
language that the interpreter is implementing. This process is not fully
automatic, but needs to be guided by the language implementer by some
source-code hints. Semi-automatically generating a JIT compiler has many advantages
over writing one manually, which is an error-prone and tedious process. The
generated JIT has the same semantics as the interpreter by construction, and all
languages implemented using PyPy benefit from improvements to the JIT generator.
The JIT that is produced by PyPy's JIT generator is a \emph{tracing JIT
compiler}, a concept which will be explained in more detail in the next section.

@@

Tracing JITs are a recently more popular approach to write just-in-time
compilers for dynamic languages \cite{XXX}. Their origins lie in the Dynamo
project, that used a tracing approach to optimize machine code using execution
traces \cite{XXX}. They were then adapted to be used for a very light-weight
Java VM \cite{XXX} and afterwards used in several implementations of dynamic
languages, such as JavaScript \cite{XXX}, Lua \cite{XXX} and now Python via
PyPy.

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter records all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning.

This trace of operations is then the basis of the generated code. The trace is
optimized in some ways, and then turned into machine code. Generating machine
code is simple, because the traces are linear and the operations are very close
to machine level. The trace corresponds to one concrete execution of a loop,
therefore the code generated from it is only one possible path through the loop.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \texttt{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \texttt{if} statement is true, because if
it isn't, the rest of the trace is not valid.

When generating machine code, every guard is be turned into a quick check to
see whether the assumption still holds. When such a guard is hit during the
execution of the machine code and the assumption does not hold, the execution of
the machine code is stopped, and interpreter continues to run from that point
on. These guards are the only mechanism to stop the execution of a trace, the
loop end condition also takes the form of a guard.
